Chemical tagging: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-15 papers/clusters.tex
zeroth draft DONE

% text shit
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\sdssiii}{\project{\acronym{SDSS-III}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\eg}{\foreign{e.\,g.}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\documentname}{\textsl{Article}}

@@

@@

\begin{abstract}
% Context:
The chemical-tagging hypothesis is that every star ought to carry---%
in its detailed chemical abundances---%
an informative signature of its molecular-cloud birth location and time.
This idea has not previously yielded scientific successes, probably because of the
noise in chemical-abundance measurements.
However, we have succeeded in improving vastly spectroscopic measurements with \thecannon\ 
(now delivering individual abundances with precisions around 0.04-dex).
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars observed by the \apogee\ spectroscopic survey
in a 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

The second issue for chemical tagging---and the one we address
here---is measurement precision (and accuracy).
The current precision on abundance measurements is not high enough
(for example, \citealt{Ting2015, martel}).
That said, it is clear that the \emph{data} are precise enough;
there is enough signal-to-noise at the relevant locations in spectrum
space to deliver high-precision tags.
Our view is that the traditional methods for making abundance
measurements just aren't mature enough:
There are issues with the physical models, and the data-analysis
methods do not make use of all of the information in the data sets.
Improving the models and exploiting the entire information
content in the data is critical if we are going to deliver useful
chemical tags.

@@

\clearpage
\section{Sample and abundances}

The details about how we select, reduce, and analyze the \apogee\ data
are given in full detail in the companion papers \citep{casey16, ness16}.
We summarize briefly here.
For the spectroscopic data, we use the \apogee\ \acronym{DR12}
(\cite{dr12}) spectroscopic data reductions for red-giant stars
($\logg < 3.9$) that are not flagged as bad in the
\apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
There are 98462 [DWH CHECK] stars that make our $\logg$ and flag cuts.

@@

DWH!!  How does \thecannon\ work here?

@@

\begin{figure}[!p]
\insanefigure{all}
\caption{Whatevsinator.\label{fig:all}}
\end{figure}

\clearpage
\section{Method and results}

Although only hints of small-scale clustering in the abundance space
are visible in \figurename~\ref{fig:all}, exploration of the data indicates that
known clusters do appear in the abundance space as over-densities.
This encourages us to look for over-densities automatically in the
abundance space and see if anything found that way would be over-dense
in phase space.
The simplest method for clustering points in $D$-dimensional space is
the \emph{k-means} algorithm (see CITE for a pedagogical introduction).
Briefly, the k-means algorithm is to find the points closest (in
$D$-space) to each of $K$ centers, and then take the mean of those
points to update the locations of the centers; iterate to convergence.
The output is the locations of the $K$ centers and the assignments of
all points to centers.
This algorithm is fast and performs well in practice in problems of
this nature.

@@

DWH: Reminder that we don't need the absolute best clustering to
demonstrate the hypothesis.

@@

All of the code for the data manipulation, clustering, and
visualization is built on numpy (CITE), scikit-learn (CITE),
and matplotlib (CITE).

\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Whatevsinator.\label{fig:M13}}
\end{figure}

\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Whatevsinator.\label{fig:Pal5}}
\end{figure}

\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Whatevsinator.\label{fig:Sgr}}
\end{figure}

\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Whatevsinator.\label{fig:halo}}
\end{figure}

\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Whatevsinator.\label{fig:disk}}
\end{figure}

@@

DWH: What have we demonstrated?

DWH: For the first time/H], [we see small-scale informative features in abundance space.

DWH: When we identify the small-scale features in abundance space,they are informative in position space.

DWH: How do we show that we aren't just getting clusters because \apogee\ has plate-to-plate calibration variations?

DWH: Chemical tagging will work.

>>>>

% text shit
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\sdssiii}{\project{\acronym{SDSS-III}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
\newcommand{\galah}{\project{\acronym{GALAH}}}
\newcommand{\thecannon}{\project{The~Cannon}}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\eg}{\foreign{e.\,g.}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}

@@

% results shit
\newcommand{\totalnumber}{98462} %% DWH CHECK

@@

\begin{abstract}
% Context:
The chemical-tagging hypothesis is that every star ought to carry---%
in its detailed chemical abundances---%
an informative signature of its molecular-cloud birth location and time.
This idea has not previously yielded scientific successes, probably because of the
noise in chemical-abundance measurements.
However, we have succeeded in substantially improving spectroscopic measurements with \thecannon\ 
(now delivering individual abundances with precisions around 0.04-dex).
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars observed by the \apogee\ spectroscopic survey
in a 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

The second issue for chemical tagging---and the one we address
here---is measurement precision (and accuracy).
The current precision on abundance measurements is not high enough
(for example, \citealt{martel, ting}).
That said, it is clear that the \emph{data} are precise enough;
there is enough signal-to-noise at the relevant locations in spectrum
space to deliver high-precision tags.
Our view is that the traditional methods for making abundance
measurements just aren't mature enough:
There are issues with the physical models, and the data-analysis
methods do not make use of all of the information in the data sets.
Improving the models and exploiting the entire information
content in the data is critical if we are going to deliver useful
chemical tags.

@@

\section{Sample and abundances}

The details about how we select, reduce, and analyze the \apogee\ data
are given in full detail in the companion papers \citep{casey16, ness16}.
We summarize briefly here.
For the spectroscopic data, we use the \apogee\ \acronym{DR12}
(\cite{dr12}) spectroscopic data reductions for red-giant stars
($\logg < 3.9$) that are not flagged as bad in the
\apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
There are \totalnumber\ stars that make our $\logg$ and flag cuts.

@@

As described in the original paper \citep{thecannon} and the companion
papers \citep{casey16, ness16}, \thecannon\ is operated in two stages.
In the first stage---the traning step---the training set of stars with
measured spectra and known labels (known parameters and abundances) is
used to establish a likelihood function for a set of coefficients that
parameterize a smooth, simple prediction of spectral pixels given
labels.
\thecannon\ finds the set of coefficients that optimizes the
training-step likelihood function.
In the implementation used in this \documentname, the spectral
prediction is quadratic in labels (including all cross-terms to second
order), and there are 17 labels ($\teff$, $\logg$, and 15 abundances
referenced to H) and 8500 spectral pixels, so there are on the order
of 1.4 million coefficients set by a convex optimization in the
training step.

In the second stage---the test step---each unlabeled star is used to
establish a single-star likelihood function for labels, holding the
coefficients fixed.
\thecannon\ finds the labels that minimize the single-star likelihood
function for each test-set star.
This optimization is not convex, but it is embarrassingly parallel.
For the results shown here, the training step takes hours and the test
step takes less than an hour on a small research cluster in Cambridge.

@@

\begin{figure}[!p]
\insanefigure{all}
\caption{The full sample of \totalnumber\ stars used in this study.
  The top six panels show six two-dimensional projections of the
  empirical abundance space distribution.
  The clustering algorithm (described in
  \sectionname~\ref{sec:method}) performs clustering in a full
  15-dimensional abundance space of elements referenced to H (not Fe),
  but plots are shown here referenced to Fe for familiarity reasons.
  The bottom three panels show stellar meta-data not used in the
  clustering but used to test whether the abundance-space clusters
  are interesting spatially.\label{fig:all}}
\end{figure}

\clearpage
\section{Method and results}\label{sec:method}

Although only hints of small-scale clustering in the abundance space
are visible in \figurename~\ref{fig:all}, exploration of the data indicates that
known clusters do appear in the abundance space as over-densities.
(In general, the collapse of 15 dimensions down to 2-d projections
will hide, smooth, or dilute any structure; there is no guarantee that
a highly featured distribution in 15-d will show features in
\emph{any} 2-d projection, let alone axis-aligned, human-selected projections.)
This encourages us to look for over-densities automatically in the
abundance space and see if anything found that way would be over-dense
in phase space.
The simplest method for clustering points in $D$-dimensional space is
the \emph{k-means} algorithm (see \citealt{bishop} for a pedagogical introduction).
Briefly, the k-means algorithm is to find the points closest (in
$D$-space) to each of $K$ centers, and then take the mean of those
points to update the locations of the centers; iterate to convergence.
The output is the locations of the $K$ centers and the assignments of
all points to centers.
This algorithm is fast and performs well in practice in problems of
this nature.

@@

DWH: EXPLICITLY GIVE THE ABUNDANCE-SPACE SCALES USED.

One note on clustering algorithm decisions is warranted:
Because this project is just a demonstration of the chemical-tagging
concept, there is no requirement that we choose the best clustering
algorithm or the method that reveals the best structures.
Indeed, it is a feature of this work that even the simplest, most
generic, least tuned clustering algorithm returns interesting
structures (as shown below).
The prospects for Milky Way science only improve as the clustering
algorithm improves.

@@

DWH \& HWR: How do we show that we aren't just getting clusters because \apogee\ has plate-to-plate calibration variations?

All of the code for the data manipulation, clustering, and
visualization is built on \project{numpy} \citep{numpy},
\project{scikit-learn} \citep{sklearn}, and \project{matplotlib}
\citep{matplotlib}.
All of the code written specifically for this project is available in
two open-source code repositories at
\url{https://github.com/andycasey/AnniesLasso} and
\url{https://github.com/davidwhogg/Platypus}.

\begin{figure}[!p]
\insanefigure{cluster_0256_0253}
\caption{Same as \figurename~\ref{fig:all} except that the entire
  sample has been made gray and the members of one of the most dense
  $K=256$ abundance-space clusters has been made black (and prominent).
  This cluster, which was identified only in abundance space (six
  projections of which are the top six panels in this \figurename),
  turns out to be dominated by the halo globular cluster
  M13.\label{fig:M13}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0034}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the disrupting globular
  cluster Palomar 5.\label{fig:Pal5}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0177}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by the Sagittarius star
  cloud.\label{fig:Sgr}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0010}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a high-velocity-dispersion
  structure in the Galaxy halo.\label{fig:halo}}
\end{figure}
\begin{figure}[!p]
\insanefigure{cluster_0256_0141}
\caption{Same as \figurename~\ref{fig:M13} but for another dense
  abundance-space cluster.
  This one turns out to be dominated by a very thin young stellar
  structure in the Galaxy disk.\label{fig:disk}}
\end{figure}

@@

We have demonstrated for the first time that, in a large-scale
spectroscopic survey of stars, stars that appear similar in
chemical-abundance properties are often associated physically in
clusters or other structures.
The reverse is well established---that is, that stars that are
associated physically in clusters are similar in chemical-abundance
properties (with a recent model-free test by \citealt{bovy}).
MKN: WHAT ELSE TO CITE HERE?
However, it was not known whether abundance-space information would be
precise enough, or \emph{measurable} precisely enough to find this
structure in surveys dominated by a large mix of stars from different
origins and ages (and there was good reason to be pessimistic;
\citealt{ting}).
We have demonstrated with a straightforward (and in no way optimal)
clustering in chemical-abundance space that it is easy to find co-eval
stellar structures.

What made this possible is substantial improvements in precision
delivered by \thecannon.
We don't have an absolute measure of the precision of the measurements,
but from looking at open and globular clusters, it appears to be on the
order of or better than 0.04~dex for the median element in the list of 15.
These improvements are discussed in detail in the companion papers
\citep{casey16, ness16} about the abundance measurements.
We believe these improvements come from a combination of factors, not
limited to (a)~improvements in the determination of pseudo-continuum,
(b)~the use of more spectral range than just unblended element windows
\citep{aspcap}, and (c)~accurate spectral predictions
(\thecannon\ delivers accurate predictions because it is fit to
observed data).
Furthermore, because abundance space is high in dimensionality, even
small improvements in abundance measurement get taken to a significant
power when thinking about information gains.

There is no sense at all in which \figurename~\ref{fig:M13},
\ref{fig:Pal5}, \ref{fig:Sgr}, \ref{fig:halo}, and \ref{fig:disk} show
a representative or complete set of features found in abundance space.
These features were hand-chosen to be obviously interesting and
interpetable.
There are many other things to be found in this data set!

On the particulars of these features, a few comments:
DWH: Comments about Sag and M13: They appear in multiple clusters.
MKN: Comments about the young disk feature and what it probably is?

The \apogee\ project shows great promise for these studies, and with
the appearance of the companion papers, we will release the
chemical-abundance measurements we used here for further study.
The future is even brighter, however: \apogee\ is expanding to 19
elements, more stars, and all-sky angular coverage with \apogee2, and
\galah\ is nearly ready to release chemical abundances on a larger set
of 29 elements.
The \apogee\ elements employed in this project include alpha
(Mg, O, Si, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
and iron peak (Fe, Ni, V, Mn) elements.
\galah\ will deliver element abundances from all the major
nucleosynthetic processes, including light proton capture elements,
alpha, odd-Z, iron-peak, light and heavy slow neutron capture, and
rapid neutron capture.
Chemical tagging capabilities are expected to grow as the
\emph{product} of the measured nucleosynthetic pathways.
